Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-09 AnalyticStructure_main.tex
update

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Give that we employ real functions both for $p$ and $s$ is real we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{p\rho}^{q\eta} = (H_{0\rho\eta}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\decay{\rho}{p};\excite{q}{\eta}] \;,\nn \\
& D_{p\rho}^{q\eta} = U[\decay{\rho}{p};\decay{\eta}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.

>>>>

\vspace{1cm}
\added[id=dal]{A proposal to reorganize the material of the following two subsection B e C in a different way}

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}

Comments on the features and the physical relevance of the two sector of the excitation spectrum

\subsection{Localized sector}

\begin{itemize}
\item It is composed by a genuine set of discrete elements. 
\item The values of the excitation energies do not depend on the computational setup (observable features) and so can be compared among different approaches. 
\item The knowledge of this sector of the excitation spectrum allows us to express only the imaginary part of (proper equation that expresses the response functional written using Sokowsky-Plemely
on the susceptibility functional). 
\item A computational setup in which delocalized degrees of freedom are excluded by design can in principle describe to this sector in a exact way. 
\end{itemize}

\subsection{Continuum sector}

\begin{itemize}
 \item Not true poles of the susceptibility (but only an effective tool to compute the spectral function).
 \item Deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and the associated spectrum exhibits an explicit dependence from the size of the domain.
 \item This sector of the excitation spectrum is needed to compute imaginary parts above IP and real part for \emph{all} the values of $\omega$. Add examples on the statical polarizability. Can we comment
 \item Make some statements on the features that a computational setup must have to guarantees that this sector provides converged results. Difference between systematic and localized basis setup. Can we comment about the needs to add ghost atoms for the evaluation of statical polarizability with gaussian basis? 
\end{itemize}

\vspace{1cm}

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this aim we introduce an explicit basis $\{\ket{s}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_s X^a_{p s}\ket{s} \;, \qq
\bra{\chi^a_p} = \sum_s\bra{s}Y^a_{p s} \;. \lb{phiandchi}
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element of the basis provides a linear systems of equations for the coefficients
$X^a_{ps}$ and $Y^a_{ps}$, namely
\begin{align}
 \sum_{s'}(H_{0ss'}-\eps_p\delta_{ss'})X^a_{ps'} + \bra{s}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{ps} \;, \nn \\
 -\sum_{s'} (H_{0 s' s}-\eps_p\delta_{s s'})Y^a_{ps'} - \bra{\psi_p}\op V'[\op E_a]\ket{s}& = \Omega_a Y^a_{ps} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{s} \equiv \ketbra{s}{\psi_p}$ and $\decay{s}{p} \equiv \ketbra{\psi_p}{s}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{ps}\excite{p}{s}X^a_{ps}+\decay{s}{p}Y^a_{ps} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{s}\op V'[\op E_a]\ket{\psi_p} = \sum_{qs'}\left(U[\decay{s}{p};\excite{q}{s'}]X^a_{qs'} + U[\decay{s}{p};\decay{s'}{q}]Y^a_{qs'}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{s} = \sum_{qs'}\left(U[\excite{p}{s};\excite{q}{s'}]X^a_{qs'} + U[\excite{p}{s};\decay{s'}{q}]Y^a_{qs'}\right) \;, \nn
\end{align}
Give that we employ real functions both for $p$ and $s$ is real we also have $\excite{p}{s}(\r,\r') = \decay{s}{p}(\r',\r)$ and the density operator is symmetric in $\r \leftrightarrow \r'$.
These conditions applied together imply that $U[\excite{p}{s};\excite{q}{s'}] = U[\decay{s}{p};\decay{s'}{q}]$, as well as $U[\decay{s}{p};\excite{q}{s'}] = U[\excite{p}{s};\decay{s'}{q}]$,
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{qs'}\mat{F_{ps}^{qs'} &  D_{ps}^{qs'}  \\
- D_{ps}^{qs'} & - F_{ps}^{qs'} }
\mat{X^a_{qs'} \\ Y^a_{qs'}} = \Omega_a \mat{X^a_{ps} \\ Y^a_{ps}}
\ee
where
\begin{align}
& F_{ps}^{qs'} = (H_{0 ss'}-\eps_p\delta_{ss'})\delta_{pq} + U[\decay{s}{p};\excite{q}{s'}] \;,\nn \\
& D_{ps}^{qs'} = U[\decay{s}{p};\decay{s'}{q}] \;.
\end{align}
This is the well-known Casida Equation in the basis of transitions for the excitation of energy $\Omega_a$. The computational reliability of the
eigenvalue $\Omega_a$ depends on the capability of the basis set $\ket{s}$ to fulfill of Eq.~\eqref{phiandchi}. Note that this may be valid only for a subset of the solutions of the eigenproblem \eqref{ExcitationMatrixEq1}.
